Interactive Curvature Tensor Visualization on Digital Surfaces\thanks{This work has been mainly funded by \textsc{DigitalSnow} ANR-11-BS02-009, \textsc{KIDICO} ANR-2010-BLAN-0205 and \textsc{PRIMES Labex} ANR-11-LABX-0063/ ANR-11-IDEX-0007 research grants.}

### 2016-01-18 latex/article.tex
fixed some orthographic mistakes in the paper

\begin{abstract}\sloppy
  Interactive visualization is a very convenient tool to explore
  complex scientific data or to explore different parameter settings
  for a given processing algorithm. In this article, we present a tool
  to efficiently explore the curvature tensor on the boundary of
  potentially large and dynamic digital objects (mean and Gaussian curvatures,
  principal curvatures, principal directions and normal vector
  field). More precisely, we combine a fully parallel pipeline on GPU
  to extract an adaptive triangulated isosurface of the digital
  object, with a curvature tensor estimation at each surface point
  based on integral invariants. Integral invariants being parametrized
  by a given ball radius, our proposal allows us to explore
  interactively different radii and thus select the appropriate scale
  at which the computation is performed and visualized.

@@

\noindent\textbf{Contributions}\quad In this work, we propose a framework to
perform interactive visualization of complex 3D digital structures combined with
%a dynamic curvature tensor estimation. More precisely, we define a fully data
a dynamic curvature tensor estimation. We define a fully data
%parallel process on GPU (Graphics Processor Unit) hardware to both efficiently
parallel process on the GPU (Graphics Processor Unit) to both efficiently
extract adaptive isosurface and compute per vertex curvature tensor using
Integral Invariants estimators.
%Hence, this system allows us to visualize
This system allow us to visualize
curvature tensor in real-time on large dynamic objects. Our approach combines a
GPU implementation of pointerless octrees to represent the data, an adaptive
viewpoint-dependent mesh extraction, and a GPU implementation of integral
invariant curvature estimators.

@@

\sloppy In our context, we consider digital shapes (any subset $Z$ of $\Z^d$)
and boundaries of digital shapes $\DigBoundary{Z}$. We denote by $\DSh$ the
Gauss digitization of $\Shape$ in a $d-$dimensional grid with grid step $h$,
\emph{i.e.}  $\DSh \EqDef \{ \vz \in \Z^d,\, h\cdot\vz\in\Shape\}$. For such
digitized set $\DigShape\EqDef\DSh$, $\Body{\DigShape}{h}$ is a subset of $\R^d$
corresponding to the union of hypercubes centered at $h\cdot \vz$ for $\vz\in Z$
with edge length $h$. By doing so, both $\partial\Shape$ and
$\partial\Body{Z}{h}$ are topological boundary of objects lying in the same
space (see Fig. \ref{fig:notations}-b). Note that the combinatorial digital
boundary $\DigBoundary{Z}$ of $Z$ made of cells in a cellular Cartesian complex
(\emph{pointels}, \emph{linels}, \emph{surfels}, \ldots), can be trivially
embedded into $\R^d$ such that it coincides with $\partial\Body{Z}{h}$.

@@

In \cite{DGCI2013}, authors have defined the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
They have proved that estimating both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information have the multigrid convergence
property \cite{DGCI2013}. For short, when the digital object becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

In \cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
They have also proved the convergence of these estimators when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}. More formally, they get:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

@@

It has been shown that the radius of the ball depends on the geometry of the
underlying shape. In \cite{DGCI2014}, authors proposed a parameter-free
estimation of the radius of the ball by analyzing the shape \wrt the local shape
geometry using maximal digital straight segments of the digital boundary. In
\cite{SMI2015}, authors analyzed in scale-space (for a range of radii) these
estimators for a given digital shape. This allows them to detect features of the
shape thanks to the behavior of estimators on singularities.

@@

In the following, we use the
%extension  of MC configurations to handle adaptive structures
algorithm
proposed by \textsc{Lengyel} \emph{et al.}~\cite{lengyel2010voxel}.
First, this approach constrains the octree
structure
%in order
to make sure that the depth difference between any
two adjacent cells is at most one.
Then, \textsc{Lengyel} \emph{et al.} introduce the concept of transition cells.
Those cells are defined to be inserted between two neighboring octree cells of different depth.
With specific MC configurations for their triangulation, a crack free mesh can
be extracted.
%Then, \textsc{Lengyel} \emph{et al.} propose
%specific MC configurations for such transition cells,  leading to a
%crack-free triangulation.

@@

When dealing with large datasets, we cannot expect real-time curvature tensor
estimation if we consider the full resolution geometry, due the the huge amount of data to process. By computing a
dynamically refined approximate curvature tensor estimation joined with an
adaptive triangulation (as discussed in Section~\ref{sec:hierarchical}), we
manage to maintain a real-time framerate by giving control over the amount of
data to process at each frame. In Figure~\ref{fig:timings}, we compare timings
(in logarithmic scale) for a triangulation that is dynamically refined according
to its distance to the camera. We measured those timings using the multiresolution regular and
the hierarchical algorithm.

>>>>

\begin{abstract}\sloppy
  Interactive visualization is a very convenient tool to explore
  complex scientific data or to try different parameter settings
  for a given processing algorithm. In this article, we present a tool
  to efficiently analyze the curvature tensor on the boundary of
  potentially large and dynamic digital objects (mean and Gaussian curvatures,
  principal curvatures, principal directions and normal vector
  field). More precisely, we combine a fully parallel pipeline on GPU
  to extract an adaptive triangulated isosurface of the digital
  object, with a curvature tensor estimation at each surface point
  based on integral invariants. Integral invariants being parametrized
  by a given ball radius, our proposal allows to explore
  interactively different radii and thus select the appropriate scale
  at which the computation is performed and visualized.

@@

\noindent\textbf{Contributions}\quad In this work, we propose a framework to
perform interactive visualization of complex 3D digital structures combined with
%a dynamic curvature tensor estimation. More precisely, we define a fully data
a dynamic curvature tensor estimation. We define a fully data
%parallel process on GPU (Graphics Processor Unit) hardware to both efficiently
parallel process on the GPU (Graphics Processor Unit) to both efficiently
extract adaptive isosurface and compute per vertex curvature tensor using
Integral Invariants estimators.
%Hence, this system allows us to visualize
This system allows us to visualize
curvature tensor in real-time on large dynamic objects. Our approach combines a
GPU implementation of pointerless octrees to represent the data, an adaptive
viewpoint-dependent mesh extraction, and a GPU implementation of integral
invariant curvature estimators.

@@

\sloppy In our context, we consider digital shapes (any subset $Z$ of $\Z^d$)
and boundaries of digital shapes $\DigBoundary{Z}$. We denote by $\DSh$ the
Gauss digitization of $\Shape$ in a $d-$dimensional grid with grid step $h$,
\emph{i.e.}  $\DSh \EqDef \{ \vz \in \Z^d,\, h\cdot\vz\in\Shape\}$. For such
digitized set $\DigShape\EqDef\DSh$, $\Body{\DigShape}{h}$ is a subset of $\R^d$
corresponding to the union of hypercubes centered at $h\cdot \vz$ for $\vz\in Z$
with edge length $h$. By doing so, both $\partial\Shape$ and
$\partial\Body{Z}{h}$ are topological boundaries of objects lying in the same
space (see Fig. \ref{fig:notations}-b). Note that the combinatorial digital
boundary $\DigBoundary{Z}$ of $Z$ made of cells in a cellular Cartesian complex
(\emph{pointels}, \emph{linels}, \emph{surfels}, \ldots), can be trivially
embedded into $\R^d$ such that it coincides with $\partial\Body{Z}{h}$.

@@

In \cite{DGCI2013}, estimators are defined based on the 2D digital curvature estimator
$\CurvH{R}$ and the 3D digital mean curvature estimator $\MeanCurvH{R}$ based on
the digital volumetric integral $\VolC(Y,h) \EqDef h^d \MCard(Y)$ (we denote such
quantity $\AreaC(Y,h)$ in dimension 2):
%% Redite
%% \Jaco{Given some point $\vx \in \R^d$, the set $Y$ is the digital intersection of $\DigShape$ with the ball of radius $R/h$ and center $\vx/h$.}
%% where $Y$ is the intersection of $\DigShape$ with a ball
%% $\Ball{R/h}{\vx/h}$ with $\MCard(Y)$, the number of digital points of
%% the intersection between a ball $\Ball{R/h}{\vx/h}$ of digital radius
%% $R/h$ and $\DigShape$ (we denote such quantity $\AreaC(Y,h)$ in
%% dimension 2):
%
\begin{Definition}
  %
  Given the Gauss digitization $Z\EqDef\DSh$ of a shape $\Shape \subset \R^2$
  (or $\R^3$ for the 3D mean curvature estimator), {\em digital curvature
  estimators} are defined for any point $\vx \in \R^2$ (or $\R^3$) as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    &\CurvH{R}(Z,\vx,h) \EqDef \frac{3\pi}{2R}
    - \frac{3\AreaC(\Ball{R/h}{\vx/h} \cap
    {Z}, h)}{R^3}\,,\label{eq:dig-curvature-estimator-k}\\
    &\MeanCurvH{R}(Z,\vx,h) \EqDef \frac{8}{3R}
    - \frac{4\VolC(\Ball{R/h}{\vx/h} \cap
    Z, h)}{\pi R^4}\,.\label{eq:dig-curvature-estimator-H}
  \end{align}
  %
  where $\Ball{R/h}{\vx/h}$ is the ball of digital radius $R/h$ centered on
  digital point $(\vx/h) \in Z$.
  %
  % \AreaC(\Ball{R/h}{\vx/h} \cap \Body{\DSh}{h}, h) \EqDef h^2 \MCard(
  % \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ and $\VolC(\Ball{R/h}{\vx/h} \cap
  % \Body{\DSh}{h}, h) \EqDef h^3 \MCard( \Ball{R/h}{\vx/h} \cap \Body{\DSh}{h} )$ are
  % the number of digital points in the intersection between $\Body{\DSh}{h}$ and
  % the ball $\Ball{R/h}{\vx/h}$ of radius $R/h$ centered on $\vx/h$, scaled by
  % $h^d$.\todo[inline]{J: Hum... pas top...}
  %
  % Similarly, $\hat{\lambda}_1$
  % and $\hat{\lambda}_2$ are the two greatest eigenvalues of the covariance
  % matrix of the digital points in the intersection between $Z$ and the ball.
  %
\end{Definition}
%
Estimators of both 2D curvature and 3D mean curvature on digital
objects with volumetric integral information are proved to be multigrid convergent
property \cite{DGCI2013}. For short, when the digital object becomes finer and
finer, \ie when the digitization step $h$ tends to zero, the estimated quantity
on $\Bd{\Body{\DSh}{h}}$ converges (theoretically and experimentally) to the
associated one on $\Bd{\Shape}$.
%
%Since our digital estimators are defined on
%digital shapes, in order to compare results with the exact curvature on
%Euclidean shape, we need to match points on boundary of the digital
%shape and points on the border of the Euclidean one.
%
Setting $R = kh^\frac{1}{3}$ on convex shapes
with at least $C^3$-boundary and bounded curvature, where $k$ is a constant
related to the maximal curvature of the shape, they have:
%
\begin{align}
\forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
\forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
&\CurvH{R}(\DSh, \vxH, h) = \Curv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,,\\
&\MeanCurvH{R}(\DSh, \vxH, h) = \MeanCurv(\Shape, \vx) + O\left(h^{\frac{1}{3}}\right)\,.
\end{align}

In \cite{CVIU2014}, authors have also defined 3D digital principal curvature
estimators $\PrincCurvH{1}{R}$ and $\PrincCurvH{2}{R}$ on $Z\subset\Z^3$ based
on digital moments:
%
\begin{Definition}
  %
  Given the Gauss digitization $Z \EqDef \DSh$ of a shape $\Shape \subset \R^3$,
  {\em 3D digital principal curvature estimators} are defined for any point $\vx
  \in \R^3$ as:
  %
  \begin{align}
    \forall 0 < h < R,\quad
    \PrincCurvH{1}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_2 - 3\hat{\lambda}_1) + \frac{8}{5 R} \,,\label{eq:dig-curvature-estimator-k1}\\
    \PrincCurvH{2}{R}(Z,\vx,h) &\EqDef \frac{6}{\pi {R}^6}(\hat{\lambda}_1 - 3\hat{\lambda}_2) + \frac{8}{5 R} \,,
    \label{eq:dig-curvature-estimator-k2}
  \end{align}
  %
  where $\hat{\lambda}_1$ and $\hat{\lambda}_2$ are the two greatest eigenvalues
  of the covariance matrix of the intersection between the digital shape
  $\DigShape$ and a ball of digital radius $R/h$ centered on digital point
  $\vx/h$.
  %
\end{Definition}
%
The covariance matrix needs to compute digital moments of order $0$, order $1$
and order $2$ (see Equation~20 of \cite{CVIU2014} for more details).
% ($\DMom{000}{h}$), of order $1$ ($\DMom{100}{h}$, $\DMom{010}{h}$ and
% $\DMom{001}{h}$) and of order $2$ ($\DMom{200}{h}$, $\DMom{020}{h}$,
% $\DMom{002}{h}$, $\DMom{110}{h}$, $\DMom{011}{h}$ and $\DMom{101}{h}$, see
% Equation~20 of \cite{CVIU2014} for more details).
These estimators are proven convergent when setting the ball
radius $h$ in $R = kh^\frac{1}{3}$, where $k$ is a constant related to the
maximal curvature of the shape, on convex shapes with at least $C^3$-boundary
and bounded curvature \cite{CVIU2014}. More formally, it leads to:
%
\begin{align}
  \forall 0 < h \le h_0,\,\, &\forall \vx \in \Bd{\Shape},\,\,
  \forall \vxH \in \Bd{\Body{\DSh}{h}} \text{~with~} \| \vxH -\vx\|_\infty \le h, \nonumber \\
  &\PrincCurvH{1}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{1}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,,\\
  &\PrincCurvH{2}{R}(\DigF{\Shape}{h},\vxH, h) = \PrincCurv{2}(\Shape,\vx) + O\left(h^{\frac{1}{3}}\right) \,.
\end{align}

@@

It has been shown that the radius of the ball depends on the geometry of the
underlying shape. In \cite{DGCI2014}, a proposal was made for a parameter-free
estimation of the radius of the ball by analyzing the shape \wrt the local shape
geometry using maximal digital straight segments of the digital boundary. In
\cite{SMI2015}, these estimators have been analyzed in scale-space (for a range of radii)  for a given digital shape. This allows to detect features of the
shape thanks to the behavior of estimators on singularities.

@@

In the following, we use the
%extension  of MC configurations to handle adaptive structures
algorithm
proposed by \textsc{Lengyel} \emph{et al.}~\cite{lengyel2010voxel}.
First, this approach forces the octree
structure
%in order
to make sure that the depth difference between any
two adjacent cells is at most one.
Then, \textsc{Lengyel} \emph{et al.} introduce the concept of transition cells.
Those cells are defined to be inserted between two neighboring octree cells of different depth.
With specific MC configurations for their triangulation, a crack free mesh can
be extracted.
%Then, \textsc{Lengyel} \emph{et al.} propose
%specific MC configurations for such transition cells,  leading to a
%crack-free triangulation.

@@

When dealing with large datasets, we cannot expect real-time curvature tensor
estimation if we consider the full resolution geometry, due to the huge amount of data to process. By computing a
dynamically refined approximate curvature tensor estimation joined with an
adaptive triangulation (as discussed in Section~\ref{sec:hierarchical}), we
manage to maintain a real-time framerate by giving control over the amount of
data to process at each frame. In Figure~\ref{fig:timings}, we compare timings
(in logarithmic scale) for a triangulation that is dynamically refined according
to its distance to the camera. We measured those timings using the multiresolution regular and
the hierarchical algorithm.
